Multiresolution Neural Field Estimation

### 2011-02-28 Multiresolution_framework/ltx/manuscript.tex
more edits to derivation
added cyan for cutting texts

\usepackage{color}
\newcommand{\mike}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}} 
\newcommand{\parham}[1]{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}} 
\newcommand{\ken}[1]{\textsf{\emph{\textbf{\textcolor{magenta}{#1}}}}}

@@

\section{Introduction}
\IEEEPARstart{T}{his} paper introduces a multiresolution data-driven framework for neural field modeling. Neural field models describe the meso and macroscopic dynamics of the human brain. Modeling the brain this level is extremely important, since diseases such as epilepsy and Parkinson's disease are manifested at this scale through pathological oscillations. The ability to create patient-specific neural field models will not only contribute to our knowledge base of poorly understood diseases such as epilepsy, but may also enable the development of new treatment strategies. This is particularly poignant with the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices using corrective or modulating electrical stimulation operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using systems theory. The design of a stimulation strategy using systems theory requires a model. Neural mass and neural field models are the ideal candidates for estimation and control algorithms. It is expected that parameters of the neural fields models will be patient specific and will therefore need to be inferred from data.

@@

\section{Integro-Difference Equation Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see \dean{cite NeuroImage paper} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f\left(.\right)$. \dean{The activation function is of the form $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, we denote the current time step by $t$ and the future time step by $t+1$. The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$, is corrupted by an i.i.d normally distributed zero-mean white noise $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$ with $\mathbf{\Sigma}_{\epsilon}=\sigma_{\epsilon}^2\mathbf I_{n_y} $. The superscript $\top$ denotes the transpose operator.

@@

The vectors \dean{(block matrices?)} of the kernel and the field scaling and wavelet functions, $\boldsymbol\lambda$ and $\boldsymbol\mu$ respectively, are defined by
\begin{align}
 \boldsymbol\lambda^\top(r-r')=&\left[ \boldsymbol\phi_{j_0}^\top(r-r') \quad \boldsymbol\psi_{j_0}^\top(r-r') \quad \boldsymbol\psi_{j_0+1}^\top(r-r') \right. \nonumber \\
&\left. \cdots \quad \boldsymbol\psi_{j}^\top(r-r')\right] \label{KernelBasisVector} \\
 \boldsymbol\mu^\top(r)=&\left[ \boldsymbol\phi_{j_0}^\top(r) \quad \boldsymbol\psi_{j_0}^\top(r) \quad \boldsymbol\psi_{j_0+1}^\top(r) \cdots \boldsymbol\psi_{j}^\top(r) \right]. 
\label{FieldBasisVector}
\end{align}
The vectors in Eqs. (\ref{KernelBasisVector}) and (\ref{FieldBasisVector}) are constructed in the same manner as the vectors in Eqs. (\ref{KernelWeights}) and (\ref{FieldWeights}).

@@

\section{B-spline wavelet IDE model}
The multiresolution approximation can be implemented using various classes of scaling and wavelet functions. B-spline functions were chosen due their compact support and the ability to analytically calculate convolution and inner product to form the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how the matrices are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
\begin{align}
N_{m}\left(r\right)=&\left(N_{m-1}\ast N_{1}\right)\left(r\right)\nonumber \\
=&\int_0^{1} N_{m-1}\left( r-r'\right)dr' \quad m>1,
\label{SplineConvolutionIntegral}
\end{align}
where $\ast$ denotes convolution and $N_1\left(s\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Eq.~\ref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ times convolution of indicator function with itself since
\begin{equation}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
\end{equation}
Using the associativity property of convolution we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of Eq.~\eqref{eq:BsplineConvolution} is (see \cite{Chui1992} for proof) 
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. The $m^{th}$ order compactly supported \dean{(is there a non-compactly supported B-spline?)} B-spline wavelet function is defined as (see \cite{Chui1992}) 
\begin{align}
 \varphi & _{m}\left(r\right) = \sum_{n=0}^{3m-2} q_n N_{m}\left(2r-n\right) \\
 q & _n = \frac{\left(-1\right)^n}{2^{m-1}} \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$},
\end{align}
where $q_n$ are the coefficients. By exploiting properties Eqs.~\ref{eq:BsplineConvolution} and \ref{eq:BsplineInnerProduct} the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta} can be computed. In order to calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$ the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale.  In this paper $4^{th}$ order cardinal B-spline scaling and wavelet functions are used. Therefore, the $8^{th}$ and $12^{th}$ order B-spline values at integer points are required (see \cite{Goswami1999}) to compute the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta}.

>>>>

\usepackage{color}
\newcommand{\mike}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}} 
\newcommand{\parham}[1]{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}} 
\newcommand{\ken}[1]{\textsf{\emph{\textbf{\textcolor{magenta}{#1}}}}} 
\newcommand{\cut}[1]{\textsf{\emph{\textbf{\textcolor{cyan}{#1}}}}}

@@

\section{Introduction}
\IEEEPARstart{T}{his} paper introduces a multiresolution data-driven framework for neural field modeling. Neural field models describe the meso and macroscopic dynamics of the human brain. Modeling the brain at this level is extremely important, since diseases such as epilepsy and \parham{Parkinson's disease} are manifested at this scale through pathological oscillations. The ability to create patient-specific neural field models will not only contribute to our knowledge base of poorly understood diseases such as epilepsy, but may also enable the development of new treatment strategies. This is particularly poignant with the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices using corrective or modulating electrical stimulation operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using systems theory. The design of a stimulation strategy using systems theory requires a model. Neural mass and neural field models are the ideal candidates for estimation and control algorithms. It is expected that parameters of the neural fields models will be patient specific and will therefore need to be inferred from data.

@@

\section{Integro-Difference Equation Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see \dean{cite NeuroImage paper} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f\left(.\right)$. \dean{The activation function is of the form $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, \cut{we denote the current time step by $t$ and the future time step by $t+1$.} The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$, is corrupted by an i.i.d normally distributed zero-mean white noise $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$ with $\mathbf{\Sigma}_{\epsilon}=\sigma_{\epsilon}^2\mathbf I_{n_y} $. The superscript $\top$ denotes the transpose operator.

@@

The vectors \dean{(block matrices?)} of the kernel and the field scaling and wavelet functions, $\boldsymbol\lambda$ and $\boldsymbol\mu$ respectively, are defined by
\begin{align}
 \boldsymbol\lambda^\top & (r-r')=\left[ \boldsymbol\phi_{j_0}^\top(r-r') \quad \boldsymbol\psi_{j_0}^\top(r-r') \quad \boldsymbol\psi_{j_0+1}^\top(r-r') \right. \nonumber \\
&\left. \cdots \quad \boldsymbol\psi_{j}^\top(r-r')\right] \label{KernelBasisVector} \\
 \boldsymbol\mu^\top & (r)=\left[ \boldsymbol\phi_{j_0}^\top(r) \quad \boldsymbol\psi_{j_0}^\top(r) \quad \boldsymbol\psi_{j_0+1}^\top(r) \cdots \boldsymbol\psi_{j}^\top(r) \right]. 
\label{FieldBasisVector}
\end{align}
The vectors in Eqs. (\ref{KernelBasisVector}) and (\ref{FieldBasisVector}) are constructed in the same manner as the vectors in Eqs. (\ref{KernelWeights}) and (\ref{FieldWeights}).

@@

% \section{B-spline wavelet IDE model}
The multiresolution approximation can be implemented using various classes of scaling and wavelet functions. B-spline functions were chosen due their compact support and the ability to analytically calculate convolution and inner product to form the matrices $ \boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how the matrices are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 
\begin{align}
N_{m}\left(r\right)=&\left(N_{m-1}\ast N_{1}\right)\left(r\right)\nonumber \\
=&\int_0^{1} N_{m-1}\left( r-r'\right)dr' \quad m>1,
\label{SplineConvolutionIntegral}
\end{align}
where $\ast$ denotes convolution and $N_1\left(s\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Eq.~\ref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ times convolution of indicator function with itself since
\begin{equation}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
\end{equation}
Using the associativity property of convolution we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of Eq.~\eqref{eq:BsplineConvolution} is (see \cite{Chui1992} for proof) 
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. The $m^{th}$ order compactly supported \dean{(is there a non-compactly supported B-spline?)} B-spline wavelet function is defined as (see \cite{Chui1992}) 
\begin{align}
 \varphi & _{m}\left(r\right) = \sum_{n=0}^{3m-2} q_n N_{m}\left(2r-n\right) \\
 q & _n = \frac{\left(-1\right)^n}{2^{m-1}} \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$},
\end{align}
where $q_n$ are the coefficients. By exploiting properties Eqs.~\ref{eq:BsplineConvolution} and \ref{eq:BsplineInnerProduct} the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta} can be computed. In order to calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$ the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale.  In this paper $4^{th}$ order cardinal B-spline scaling and wavelet functions are used. Therefore, the $8^{th}$ and $12^{th}$ order B-spline values at integer points are required (see \cite{Goswami1999}) to compute the integrals in Eqs.~\ref{eq:Lambdax} and \ref{eq:Lambdatheta}.
